---
title: "Introduction to Single Cell RNA-Seq Part 6: Enrichment and model-based differential expression"
author: "Bioinformatics Core"
date: "`r Sys.Date()`"
output:
    html_document:
      keep_md: TRUE
      toc: TRUE
---

# Introduction to Single Cell RNA-Seq Part 6: Enrichment and model-based differential expression
```{r setup_opt, include=FALSE}
knitr::opts_chunk$set(width = 450, error = FALSE, warning = FALSE, message = FALSE)
```

## Set up workspace
```{r workspace}
library(Seurat)
library(limma)
library(topGO)
library(dplyr)
library(kableExtra)
set.seed(12345)
experiment.aggregate <- readRDS("scRNA_workshop-05.rds")
Idents(experiment.aggregate) <- "finalcluster"
```

## 1. Gene Ontology (GO) Enrichment of Genes Expressed in a Cluster
[Gene Ontology](http://geneontology.org/docs/ontology-documentation/) provides a controlled vocabulary for describing gene products.  Here we use enrichment analysis to identify GO terms that are over-represented among the gene expressed in cells in a given cluster. 
```{r GO}
cluster10 <- subset(experiment.aggregate, idents = '10')
expr <- as.matrix(GetAssayData(cluster10))

# Select genes that are expressed > 0 in at least half of cells
n.gt.0 <- apply(expr, 1, function(x)length(which(x > 0)))
expressed.genes <- rownames(expr)[which(n.gt.0/ncol(expr) >= 0.5)]
all.genes <- rownames(expr)

# define geneList as 1 if gene is in expressed.genes, 0 otherwise
geneList <- ifelse(all.genes %in% expressed.genes, 1, 0)
names(geneList) <- all.genes

# Create topGOdata object
	GOdata <- new("topGOdata",
		ontology = "BP", # use biological process ontology
		allGenes = geneList,
		geneSelectionFun = function(x)(x == 1),
              annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "symbol")
# Test for enrichment using Fisher's Exact Test
	resultFisher <- runTest(GOdata, algorithm = "elim", statistic = "fisher")
	GenTable(GOdata, Fisher = resultFisher, topNodes = 20, numChar = 60) %>%
	  kable() %>%
	  kable_styling("striped", fixed_thead = TRUE)
```
* Annotated: number of genes (out of all.genes) that are annotated with that GO term
* Significant: number of genes that are annotated with that GO term and meet our criteria for "expressed"
* Expected: Under random chance, number of genes that would be expected to be annotated with that GO term and meeting our criteria for "expressed"
* Fisher: (Raw) p-value from Fisher's Exact Test

## 2. Model-based DE analysis in limma
[limma](https://bioconductor.org/packages/release/bioc/html/limma.html) is an R package for differential expression analysis of bulk RNASeq and microarray data.  We apply it here to single cell data.

Limma can be used to fit any linear model to expression data and is useful for analyses that go beyond two-group comparisons.  A detailed tutorial of model specification in limma is available [here](https://ucdavis-bioinformatics-training.github.io/2021-June-RNA-Seq-Analysis/data_analysis/DE_Analysis_mm_with_quizzes) and in the [limma User's Guide](https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf).

```{r DE}
# filter genes to those expressed in at least 10% of cells
keep <- rownames(expr)[which(n.gt.0/ncol(expr) >= 0.1)]
expr2 <- expr[keep,]

# Set up "design matrix" with statistical model
cluster10$proper.group <- make.names(cluster10$group)
mm <- model.matrix(~0 + proper.group + S.Score + G2M.Score + percent_MT + nFeature_RNA, data = cluster10[[]])
head(mm) %>%
	  kable() %>%
	  kable_styling("striped", fixed_thead = TRUE)
tail(mm) %>%
	  kable() %>%
	  kable_styling("striped", fixed_thead = TRUE)

# Fit model in limma
fit <- lmFit(expr2, mm)
head(coef(fit)) %>%
	  kable() %>%
	  kable_styling("striped", fixed_thead = TRUE)

# Test 'Normal' - 'Colorectal.Cancer'
contr <- makeContrasts(proper.groupNormal - proper.groupColorectal.Cancer, levels = colnames(coef(fit)))
contr %>%
	  kable() %>%
	  kable_styling("striped", fixed_thead = TRUE)
fit2 <- contrasts.fit(fit, contrasts = contr)
fit2 <- eBayes(fit2)
out <- topTable(fit2, n = Inf, sort.by = "P")
head(out, 30) %>%
	  kable() %>%
	  kable_styling("striped", fixed_thead = TRUE)
```

**Output columns:**

* logFC: log fold change (since we are working with Seurat's natural log transformed data, will be natural log fold change)
* AveExpr: Average expression across all cells in expr2
* t: logFC divided by its standard error
* P.Value: Raw p-value (based on t) from test that logFC differs from 0
* adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value
* B: log-odds that gene is DE 

## Save files
```{r write.csv}
write.csv(GenTable(GOdata, Fisher = resultFisher), file = "cluster10_GOdata.csv")
write.csv(out, file = "cluster10_Normal-Colorectal.Cancer_topTable.csv")
```

## Prepare for the next section

#### Download Rmd document
```{r download_Rmd, eval=F}
download.file("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2023-December-Single-Cell-RNA-Seq-Analysis/main/data_analysis/07-doublet_detection.Rmd", "07-doublet_detection.Rmd")
```

#### Session Information
```{r sessionInfo, warning=FALSE,error=FALSE,message=FALSE}
sessionInfo()
```
