---
title: "Single Cell RNAseq Part 5"
author: "Bioinformatics Core"
output:
    html_document:
      keep_md: TRUE
---

<script>
function buildQuiz(myq, qc){
  // variable to store the HTML output
  const output = [];

  // for each question...
  myq.forEach(
    (currentQuestion, questionNumber) => {

      // variable to store the list of possible answers
      const answers = [];

      // and for each available answer...
      for(letter in currentQuestion.answers){

        // ...add an HTML radio button
        answers.push(
          `<label>
            <input type="radio" name="question${questionNumber}" value="${letter}">
            ${letter} :
            ${currentQuestion.answers[letter]}
          </label><br/>`
        );
      }

      // add this question and its answers to the output
      output.push(
        `<div class="question"> ${currentQuestion.question} </div>
        <div class="answers"> ${answers.join('')} </div><br/>`
      );
    }
  );

  // finally combine our output list into one string of HTML and put it on the page
  qc.innerHTML = output.join('');
}

function showResults(myq, qc, rc){

  // gather answer containers from our quiz
  const answerContainers = qc.querySelectorAll('.answers');

  // keep track of user's answers
  let numCorrect = 0;

  // for each question...
  myq.forEach( (currentQuestion, questionNumber) => {

    // find selected answer
    const answerContainer = answerContainers[questionNumber];
    const selector = `input[name=question${questionNumber}]:checked`;
    const userAnswer = (answerContainer.querySelector(selector) || {}).value;

    // if answer is correct
    if(userAnswer === currentQuestion.correctAnswer){
      // add to the number of correct answers
      numCorrect++;

      // color the answers green
      answerContainers[questionNumber].style.color = 'lightgreen';
    }
    // if answer is wrong or blank
    else{
      // color the answers red
      answerContainers[questionNumber].style.color = 'red';
    }
  });

  // show number of correct answers out of total
  rc.innerHTML = `${numCorrect} out of ${myq.length}`;
}
</script>

# Part 5: Enrichment, Model-Based DE, and Cell-Type Identification

```{r setup_opt, warning=FALSE,error=FALSE,message=FALSE, echo = F}
options(width = 450)
```

## Load libraries
```{r load_library, warning=FALSE,error=FALSE,message=FALSE}
library(Seurat)
library(limma)
library(topGO)
library(HGNChelper)
library(dplyr)
```

## Load the Seurat object
```{r load_rdata, warning=FALSE,error=FALSE,message=FALSE}
experiment.merged <- readRDS("scRNA_workshop_4.rds")
experiment.merged
Idents(experiment.merged) <- "finalcluster"
```

# 1. Gene Ontology (GO) Enrichment of Genes Expressed in a Cluster
[Gene Ontology](http://geneontology.org/docs/ontology-documentation/) provides a controlled vocabulary for describing gene products.  Here we use enrichment analysis to identify GO terms that are overrepresented among the gene expressed in cells in a given cluster. 

```{r GO, warning=FALSE,error=FALSE,message=FALSE}
cluster10 <- subset(experiment.merged, idents = '10')
expr <- as.matrix(GetAssayData(cluster10))

# Select genes that are expressed > 0 in at least half of cells
n.gt.0 <- apply(expr, 1, function(x)length(which(x > 0)))
expressed.genes <- rownames(expr)[which(n.gt.0/ncol(expr) >= 0.5)]
all.genes <- rownames(expr)

# define geneList as 1 if gene is in expressed.genes, 0 otherwise
geneList <- ifelse(all.genes %in% expressed.genes, 1, 0)
names(geneList) <- all.genes

# Create topGOdata object
	GOdata <- new("topGOdata",
		ontology = "BP", # use biological process ontology
		allGenes = geneList,
		geneSelectionFun = function(x)(x == 1),
              annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "symbol")
# Test for enrichment using Fisher's Exact Test
	resultFisher <- runTest(GOdata, algorithm = "elim", statistic = "fisher")
	GenTable(GOdata, Fisher = resultFisher, topNodes = 20, numChar = 60)
```
* Annotated: number of genes (out of all.genes) that are annotated with that GO term
* Significant: number of genes that are annotated with that GO term and meet our criteria for "expressed"
* Expected: Under random chance, number of genes that would be expected to be annotated with that GO term and meeting our criteria for "expressed"
* Fisher: (Raw) p-value from Fisher's Exact Test

## Quiz 1

<div id="quiz1" class="quiz"></div>
<button id="submit1">Submit Quiz</button>
<div id="results1" class="output"></div>
<script>
quizContainer1 = document.getElementById('quiz1');
resultsContainer1 = document.getElementById('results1');
submitButton1 = document.getElementById('submit1');

myQuestions1 = [
  {
    question: "What GO term is most significantly enriched for genes expressed in cluster 10?",
    answers: {
      a: "intracellular chloride ion homeostasis",
      b: "cytoplasmic translation",
      c: "protein folding",
      d: "ribosomal large subunit biogenesis"
    },
    correctAnswer: "a"
  },
  {
    question: "How many genes annotated with the top GO term are expressed in cluster 10?",
    answers: {
      a: "8",
      b: "0.32",
      c: "0",
      d: "4"
    },
    correctAnswer: "d"
  },
  {
    question: "How many expressed genes would be expected to be annotated with the top GO term under random chance?",
    answers: {
      a: "8",
      b: "0.32",
      c: "0",
      d: "4"
    },
    correctAnswer: "b"
  }
];

buildQuiz(myQuestions1, quizContainer1);
submitButton1.addEventListener('click', function() {showResults(myQuestions1, quizContainer1, resultsContainer1);});
</script>

# 2. Model-based DE analysis in limma
[limma](https://bioconductor.org/packages/release/bioc/html/limma.html) is an R package for differential expression analysis of bulk RNASeq and microarray data.  We apply it here to single cell data.

Limma can be used to fit any linear model to expression data and is useful for analyses that go beyond two-group comparisons.  A detailed tutorial of model specification in limma is available [here](https://ucdavis-bioinformatics-training.github.io/2021-June-RNA-Seq-Analysis/data_analysis/DE_Analysis_mm_with_quizzes) and in the [limma User's Guide](https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf).

```{r}
# filter genes to those expressed in at least 10% of cells
keep <- rownames(expr)[which(n.gt.0/ncol(expr) >= 0.1)]
expr2 <- expr[keep,]

# Set up "design matrix" with statistical model
cluster10$proper.ident <- make.names(cluster10$orig.ident)
mm <- model.matrix(~0 + proper.ident + S.Score + G2M.Score + percent_MT + nFeature_RNA, data = cluster10[[]])
head(mm)
tail(mm)

# Fit model in limma
fit <- lmFit(expr2, mm)
head(coef(fit))

# Test 'B001-A-301' - 'A001-C-007'
contr <- makeContrasts(proper.identB001.A.301 - proper.identA001.C.007, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, contrasts = contr)
fit2 <- eBayes(fit2)
out <- topTable(fit2, n = Inf, sort.by = "P")
head(out, 30)
```

### Output columns:
* logFC: log fold change (since we are working with Seurat's natural log transformed data, will be natural log fold change)
* AveExpr: Average expression across all cells in expr2
* t: logFC divided by its standard error
* P.Value: Raw p-value (based on t) from test that logFC differs from 0
* adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value
* B: log-odds that gene is DE 

## Quiz 2

<div id="quiz2" class="quiz"></div>
<button id="submit2">Submit Quiz</button>
<div id="results2" class="output"></div>
<script>
quizContainer2 = document.getElementById('quiz2');
resultsContainer2 = document.getElementById('results2');
submitButton2 = document.getElementById('submit2');

myQuestions2 = [
  {
    question: "How many genes have adj.P.Val < 0.05?",
    answers: {
      a: "457",
      b: "131",
      c: "0",
      d: "100"
    },
    correctAnswer: "a"
  },
  {
    question: "How many genes are significantly (adj.P.Val < 0.05) downregulated in B001-A-301 relative to A001-C-007?",
    answers: {
      a: "273",
      b: "65",
      c: "0",
      d: "24"
    },
    correctAnswer: "a"
  },
  {
    question: "Revise the code to test 'A001-C-007' - 'A001-C-104'.  How many genes are differentially expressed between these groups? (adj.P.Val < 0.05)?  (Hint: ?makeContrasts)",
    answers: {
      a: "0",
      b: "21",
      c: "283",
      d: "27"
    },
    correctAnswer: "b"
  }
];

buildQuiz(myQuestions2, quizContainer2);
submitButton2.addEventListener('click', function() {showResults(myQuestions2, quizContainer2, resultsContainer2);});
</script>

# Cell type identification with ScType
[ScType](https://www.nature.com/articles/s41467-022-28803-w) assigns our clusters from Seurat to a cell type based on a hierarchical external database.

The database supplied with the package is human but users can supply their own data.  

More details are available on [Github](https://github.com/IanevskiAleksandr/sc-type):

Source ScType functions from Github:
```{r}
# load gene set preparation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
# load cell type annotation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")
```

Read in marker database:
```{r, warning = FALSE}
# DB file
db_ = "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_full.xlsx";
tissue = "Intestine" # e.g. Immune system,Pancreas,Liver,Eye,Kidney,Brain,Lung,Adrenal,Heart,Intestine,Muscle,Placenta,Spleen,Stomach,Thymus 

# prepare gene sets
gs_list = gene_sets_prepare(db_, tissue)
```

Let's take a look at the structure of the marker database:
```{r}
head(gs_list)
```

Let's add a cell type and associated markers (markers from https://panglaodb.se/markers.html?cell_type=%27Tuft%20cells%27#google_vignette):
```{r}
gs_list$gs_positive$`Tuft cells` <- c("SUCNR1", "FABP1", "POU2F3", "SIGLECF", "CDHR2", "AVIL", "ESPN", "LRMP", "TRPM5", "DCLK1", "TAS1R3", "SOX9", "TUBB5", "CAMK2B", "GNAT3", "IL25", "PLCB2", "GFI1B", "ATOH1", "CD24A", "ASIC5", "KLF3", "KLF6", "DRD3", "NRADD", "GNG13", "NREP", "RGS2", "RAC2", "PTGS1", "IRF7", "FFAR3", "ALOX5", "TSLP", "IL4RA", "IL13RA1", "IL17RB", "PTPRC")
gs_list$gs_negative$`Tuft cells` <- NULL
```

```{r}
# get cell-type by cell matrix
scale.data <- GetAssayData(experiment.merged, "scale")
es.max = sctype_score(scRNAseqData = scale.data, scaled = TRUE, 
                      gs = gs_list$gs_positive, gs2 = gs_list$gs_negative) 
# cell type scores for first cell
es.max[,1]
```

We will derive a cluster-level score by summing the cell level scores within each cluster
```{r}
tmp <- lapply(sort(unique(experiment.merged$finalcluster)), function(cl, nkeep = 3){
  es.max.cl <- sort(rowSums(es.max[, experiment.merged$finalcluster == cl]), decreasing = TRUE)
  out <- head(data.frame(finalcluster = cl, ScType = names(es.max.cl), scores = es.max.cl, ncells = sum(experiment.merged$finalcluster == cl)), nkeep)
  out$rank <- 1:nkeep
  return(out)
})
cL_results.top <- do.call("rbind", tmp)
cL_results.top1 <- subset(cL_results.top, rank == 1)
```

Merge ScType cluster-level results with Seurat object
```{r}
tmp <- data.frame(cell = colnames(experiment.merged), finalcluster = experiment.merged$finalcluster)
tmp <- left_join(tmp, cL_results.top1, by = "finalcluster")
experiment.merged$ScType <- tmp$ScType
```

```{r}
DimPlot(experiment.merged, group.by = "ScType")
```

The ScType developer suggests that assignments with a score less than (number of cells in cluster)/4 are low confidence and should be set to unknown:
```{r}
tmp <- data.frame(cell = colnames(experiment.merged), finalcluster = experiment.merged$finalcluster)
tmp <- left_join(tmp, cL_results.top1, by = "finalcluster")
# set assignments with scores less than ncells/4 to unknown
tmp$ScType.filtered <- ifelse(tmp$scores < tmp$ncells/4, "Unknown", tmp$ScType)
experiment.merged$ScType.filtered <- tmp$ScType.filtered
DimPlot(experiment.merged, group.by = "ScType.filtered")
```

## Get the next Rmd file
```{r getnextfile, eval=TRUE}
download.file("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2022-December-Single-Cell-RNA-Seq-Analysis/main/data_analysis/scRNA_Workshop-PART6.Rmd", "scRNA_Workshop-PART6.Rmd")
```

                 
## Session Information
```{r session_info, warning=FALSE,error=FALSE,message=FALSE}
sessionInfo()
```
